Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-09 paper.tex
major updating

\begin{abstract}
  Baroclinic astrophysical disks with equilibrium rotation profiles
  $\Omega$ that depends on the distance away from the disk midplane are
  potentially unstable to the Vertical Shear Instability (VSI). It is
  is important to understand the linear instability before
  considering the VSI as a hydrodynamic route to turbulent angular
  momentum transport in accretion disks. We study the linear,
  axisymmetric VSI in radially-local but vertically-global models
  of accretion disks. We include an energy equation with thermal
  relaxation on a characteristic timescale $t_c=\beta\Omega_k^{-1}$,
  where $\beta$ is a dimensionless parameter and $\Omega_k$ is the
  Keplerian rotation frequency, and explore how the VSI depends on the
  thermodynamic response  of the disk by varying $\beta$. We find
  increasing $\beta$ away from zero dramatically reduces VSI growth
  rates because this generally introduces a strongly stabilizing
  vertical entropy gradient. Our linear calculations are consistent
  with previous hydrodynamic simulations. These simulations indicate that the VSI 
  requires a short thermal relaxation timscale. For vertically 
  isothermal disks, we quantify `fast'
  thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q$ is the radial power-law
  index for the equilibrium temperature profile (which gives rise to vertical 
  shear) and $\gamma$ is the adiabatic index. We apply our linear
  results to a disk model of the Minimum Mass Solar Nebulae (MMSN), and
  find the MMSN is unstable to the VSI at tens of AU with a
  characteristic growth times of tens of local orbital periods. 
\end{abstract}

>>>>

\begin{abstract}
  Baroclinic astrophysical disks with equilibrium rotation profiles
  $\Omega$ that depends on the distance away from the disk midplane are
  potentially unstable to the Vertical Shear Instability (VSI). It is
  is important to understand the linear instability before
  considering the VSI as a hydrodynamic route to turbulent angular
  momentum transport in accretion disks. We study the linear,
  axisymmetric VSI in radially-local but vertically-global models
  of accretion disks. We include an energy equation with thermal
  relaxation on a characteristic timescale $t_c=\beta\Omega_k^{-1}$,
  where $\beta$ is a dimensionless parameter and $\Omega_k$ is the
  Keplerian rotation frequency, and explore how the VSI depends on the
  thermodynamic response  of the disk by varying $\beta$. We find
  increasing $\beta$ away from zero dramatically reduces VSI growth
  rates because this generally introduces a strongly stabilizing
  vertical entropy gradient. Our linear calculations are consistent
  with previous hydrodynamic simulations, which indicates the VSI 
  requires a short thermal relaxation timscale. For vertically  
  isothermal disks, we quantify `fast'
  thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q$ is the radial power-law
  index for the equilibrium temperature profile (which gives rise to vertical 
  shear) and $\gamma$ is the adiabatic index. We apply our linear
  results to a disk model of the Minimum Mass Solar Nebulae (MMSN), and
  find the MMSN is unstable to the VSI at tens of AU with a
  characteristic growth times of tens of local orbital periods. We 
  suggest the VSI is a viable mechanism for angular momentum
  transport in the outer parts of protoplanetary disks. 
\end{abstract}

### 2015-02-09 setup.tex
major updating

In order to emphasize on the radially-local approximation, we
introduce the local Cartesian co-ordinate $(x,y,z)$ corresponding to
the the local $(r,\phi,z)$ directions, respectively. Then writing
$\delta P /\rho \equiv W$ and $c_s^2\delta\rho/\rho\equiv Q$, the
linearized system of equations are 
\begin{align}
  \ii\sigma \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \hat{g}_c
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\sigma \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \sigma\delta v_y &= r\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\sigma\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \sigma W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x.\label{lin_energy}
  % -&\ii\sigma W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\sigma\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}
It is understood that the quantities $\rho,\,c_s^2,\,\Omega,\kappa,\, 
\p_z\Omega,\, \p_r\ln{\rho} $ and $ \p_r P$ are evaluated at $r=r_0$.   
Eq. \ref{lin_mass}---\ref{lin_vz} will form the basis of our numerical
study of the vertical shear instability, but for analytical discussion
we will make further simplifications.

>>>>

In order to emphasize on the radially-local approximation, we
introduce the local Cartesian co-ordinate $(x,y,z)$ corresponding to
the the local $(r,\phi,z)$ directions, respectively. Then writing
$\delta P /\rho \equiv W$ and $c_s^2\delta\rho/\rho\equiv Q$, the
linearized system of equations are 
\begin{align}
  \ii\sigma \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \hat{g}_c
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\sigma \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \sigma\delta v_y &= r\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\sigma\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \sigma W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x.\label{lin_energy}
  % -&\ii\sigma W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\sigma\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}
It is understood that the quantities $\rho,\,c_s^2,\,\Omega,\kappa,\, 
\p_z\Omega,\, \p_r\ln{\rho} $ and $ \p_r P$ are evaluated at $r=r_0$.   
Eq. \ref{lin_mass}---\ref{lin_energy} will form the basis of our numerical
study of the vertical shear instability, but for analytical discussion
we will make further simplifications.

### 2015-02-09 summary.tex
major updating

\section{Summary and discussion}
%what have we done, relation to previous work 
In this paper we have studied the linear axisymmetric vertical shear
instability (VSI) in astrophysical disks that arises from the
height-dependence of the equilibrium angular velocity profile
$\Omega(r,z)$ of a baroclinic  disk. The VSI was analysed and
explained earlier in the local limit by \cite{urpin98} and
\cite{urpin03}, simulated by \cite{arlt04} and more  recently by
\cite{nelson13}.

@@

In accordance with the above studies, which show that the VSI   
applies to perturbations with small radial lengthscales, we retained
the radially-local approximation by using the `vertically-global
shearing box' (VGSB) formalism developed by \cite{mcnally14}, who 
showed the linear VSI studied by \cite{nelson13} in a locally
isothermal disk can be captured by the VGSB.

% The presence of a
% vertical shear, but 
% is not consistent with the
% global

We added an energy source term in the VGSB equations which relaxes the disk
temperature to its initial value on a timescale
$t_c=\beta\Omega_k^{-1}$. This form of thermal relaxation was also 
employed in the global numerical simulations of \cite{nelson13}, and allows
the thermodynamic response of the disk to be controlled.

We mostly focused on vertically isothermal disks ($\Gamma=1$) because
it permits a largely analytical discussion. Our analysis is an
extension of the linear models considered in \cite{lubow93} by
adding vertical shear, but in the limit of low-frequency
perturbations in nearly-Keplerian disks. These approximations were
made for simplicity, but also motivated by previous work
\citep{nelson13}.

For isothermal perturbations ($\gamma=1$ or $\beta=0$) we explicitly
demonstrated instability  associated with vertical shear, and that the
VSI growth rate is limited by the maximum vertical shear within the
vertical domain of interest. The fundamental VSI mode is characterized
by one node in the density perturbation at the mid-plane
(i.e. anti-symmetric about $z=0$), which corresponds to a vertical
velocity perturbation  $\delta v_z\sim\mathrm{constant}$ near $z=0$.

@@

For non-isothermal  perturbations with thermal relaxation ($\gamma>1$,
$\beta>0$) , the general dispersion relation (Eq. \ref{relax_disp}) is
complicated and requires a numerical treatment. However, for the
fundamental VSI some simplification is  possible. In particular, we quantified the critical thermal relaxation
timescale $\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$, below
which small-scale perturbations ($\khat\gg 1$) may be regarded as isothermal
and the fundamental VSI operates. We believe this explicit expression for
$\beta_\mathrm{crit}$ is the most useful result from our
investigation.

We also solved the linear stability problem numerically using a
pseudo-spectral approach. Numerical solutions are mostly consistent
with our analytical expectations, except for perturbations with small
radial wavenumber ($\khat\to0$) where the low-frequency approximation
is inappropriate. The local model is also not valid for such modes.

We briefly studied vertically non-isothermal disks using our linear 
code. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI operates and is insensitive to
thermal relaxation. When a stable stratification is introduced, the
rapid reduction in VSI growth rates by a finite thermal relaxation timescale
is qualitatively similar to that observed for vertically isothermal
disks.

@@

We applied our linear results to assess the applicability of the VSI in
protoplanetary disks, specifically the Minimum Mass Solar Nebulae (MMSN),
using disk models described in \cite{chiang10}. We find the MMSN is
unstable to the fundamental VSI at $10$s of AU, with characteristic
growth times of $10$s of orbits. However, we note the strong
temperature dependence of the physical thermal relaxation timescale
in the disk model ($\beta$ decreases as $T^{-5}$ for perturbations with small
radial scale). Thus, increasing the disk temperature even slightly can signifcantly 
shorten the typical growth times for the VSI. This is also attributed
to a larger value of $\beta_\mathrm{crit}$, since increasing $T$ also
increases the aspect-ratio $\epsilon$.

@@

\subsection{Future extensions} 
We highlight below some improvements to the present  
linear models that should be pursued:

@@

\emph{Geometry.} Extrapolating our results from the shearing box to a global disk  
has an inconsistency. Namely, the equilibrium temperature in the
box only depends on height (if at all). However, it is precisely the
radial temperature (or entropy) gradient in the global disk that
gives rise to vertical shear. We do not expect this to be a serious issue for 
perturbations with small radial wavelengths, 
%explicitly checked by including global terms 
but will need to be resolved by adopting global disk models  
if one wishes to generalize the VSI to large-scale perturbations.

>>>>

\section{Summary and discussion}
%what have we done, relation to previous work 
In this paper we have studied the linear axisymmetric vertical shear
instability (VSI) in astrophysical disks that arises from the
height-dependence of the equilibrium angular velocity profile
$\Omega(r,z)$ of a baroclinic disk. The VSI was analysed and
explained earlier in the local limit by \cite{urpin98} and
\cite{urpin03}; and simulated by \cite{arlt04} and more recently by
\cite{nelson13}.

@@

We added a source term in the energy equation which relaxes the disk
temperature to its initial value on a timescale $t_c=\beta\Omega_k^{-1}$. This
form of thermal relaxation was also employed in the global numerical
simulations of \cite{nelson13}, and allows the thermodynamic response
of the disk to be controlled.

In accordance with the above studies, which show that the VSI applies
to perturbations with small radial lengthscales, we applied the 
radially-local approximation in the linearized fluid equations. For
analytic discussion only, we further simplied the linear problem by
ignoring the background radial disk structure. This is equivalent to adopting
the `vertically-global shearing box' (VGSB) formalism developed by
\cite{mcnally14}, who showed the linear VSI studied by
\cite{nelson13} in a locally isothermal disk can be captured by the
VGSB. We find the VGSB framework is appropriate for $\beta\lesssim1$. 
This is convenient because previous works show that rapid thermal
relaxation ($\beta\ll1$) is generally required for the VSI. However,
the VGSB framework is not valid in the adiabatic limit.

%OK framework for the regime of interest 
% The presence of a
% vertical shear, but 
% is not consistent with the
% global

We mostly focused on vertically isothermal disks ($\Gamma=1$) because
it permits a largely analytical discussion. Our analysis is an
extension of the linear models considered in \cite{lubow93} by
adding vertical shear and (rapid) thermal relaxation, but in the limit of low-frequency
perturbations in nearly-Keplerian disks. These approximations were
made for simplicity, but also motivated by previous work
\citep{nelson13}.

For isothermal perturbations ($\beta=0$) we explicitly
demonstrated instability  associated with vertical shear, and that the 
VSI growth rate is limited by the maximum vertical shear within the 
vertical domain of interest. The fundamental VSI mode is characterized
by one node in the density perturbation at the mid-plane
(i.e. anti-symmetric about $z=0$), which corresponds to a vertical
velocity perturbation  $\delta v_z\sim\mathrm{constant}$ near
$z=0$.

@@

For non-isothermal  perturbations with thermal relaxation ($\gamma>1$,
$\beta>0$) , the general dispersion relation (Eq. \ref{relax_disp}) is
complicated and requires a numerical treatment. However, for the 
fundamental VSI some simplification is  possible. In particular, we
quantified the critical thermal relaxation timescale
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$, below which small-scale
perturbations may be regarded as isothermal and the fundamental VSI
operates.  We expect $\beta<\beta_\mathrm{crit}$ to be a sufficient
condition for the VSI to operate. We believe this explicit expression for
$\beta_\mathrm{crit}$ is the most useful result from our
investigation.

We also solved the linear stability problem numerically using a
pseudo-spectral approach. For numerical calculations we accounted for
the background radial disk structure for consistency. We focused
on the fundamental mode to compare with our analytical
discussion, and because \cite{nelson13} found that this mode 
eventually dominate their simualtions. Our numerical solutions are
mostly consistent with our analytic expectations, and we verified the
dependence of $\beta_\mathrm{crit}$ on disk parameters.

% as given
% by Eq. \ref{iso_vsi_cond}, provides a good measure of the required
% th 
%  except for perturbations with small radial
% wavenumber ($\khat\to0$) where the low-frequency approximation is
% inappropriate. The radially local model is also not valid for such modes.

We briefly studied vertically non-isothermal disks using our linear 
code. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI operates and is insensitive to
thermal relaxation. When a stable stratification is introduced, the
rapid reduction in VSI growth rates by a finite thermal relaxation
timescale is qualitatively similar to that observed for vertically
isothermal disks. Based on numerical results, we conjecture that the
critical thermal relaxation timescale is given by
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-\Gamma)$ in verticlaly
non-isothermal disks.

@@

We applied our linear results to assess the applicability of the VSI in
protoplanetary disks, specifically the Minimum Mass Solar Nebulae (MMSN),
using disk models described in \cite{chiang10}. We find the MMSN is
unstable to the fundamental VSI at $10$s of AU, with characteristic
growth times of $10$s of orbits. However, we note the strong
temperature dependence of the physical thermal relaxation timescale
in the disk model ($\beta$ decreases as $T^{-5}$ for perturbations with small
radial scale). Thus, increasing the disk temperature even slightly can signifcantly 
shorten the typical growth times for the VSI. This is also attributed
to a larger value of $\beta_\mathrm{crit}\propto\epsilon$, since increasing $T$ also
increases the aspect-ratio $\epsilon$.

@@

\subsection{Caveats and outlooks} 
We highlight below some improvements to the present  
linear models that should be pursued:

@@

\emph{Geometry.} An important caveat

% Extrapolating our results from the shearing box to a global disk  
% has an inconsistency. Namely, the equilibrium temperature in the
% box only depends on height (if at all). However, it is precisely the
% radial temperature (or entropy) gradient in the global disk that
% gives rise to vertical shear. We do not expect this to be a serious issue for 
% perturbations with small radial wavelengths, 
% but will need to be resolved by adopting global disk models  
% if one wishes to generalize the VSI to large-scale perturbations.
